---
title: "Seneca_drums2"
author: "Erin Hassett"
date: "2025-11-10"
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)

setwd("C:/Erin/Seneca_Drums") #change to your wd
#methane data from LiCOR
pocks<- readxl::read_xlsx("ProcessedConcs2.xlsx")
pocks<- pocks[-1,]
```
# Reading in Picarro Data
```{r}
#data from the Picarro (isotopes and gas)
df_picarro <- readxl::read_xlsx("SenecaDrums_IsotopeRatioData.xlsx",sheet = "DataComp")
df_picarro$Dilution[is.na(df_picarro$Dilution)] <- 1 # setting non-diluted samples to dilution factor of 1

# Selecting for HR or HP methane depending on measured concentration (>10ppm = HR) and cleaning up columns
df_picarro <- df_picarro %>%
  mutate(CH4.Conc = if_else(HR.CH4 > 10,HR.CH4/Dilution,HP.CH4/Dilution),
         CH4.std = if_else(HR.CH4 > 10,HR.CH4.std,HP.CH4.std),
         d13C.CH4 = if_else(HR.CH4 > 10, HR.d13C,HP.d13C),
         d13C.CH4.std = if_else(HR.CH4 > 10, HR.d13C.std,HP.d13C.std)) %>%
  dplyr::select(-c(4,5,6,7,10,11,12,13)) %>%
  drop_na(CH4.Conc)

# Separating full Name into components
df_picarro <- df_picarro %>%
  separate(
    col = Name,
    into = c("Pockmark", "in.out", "Site","rep"),
    sep = "_",
    remove = FALSE)
```
##renaming duplicate Names
```{r}
#the dilutions cause issues with having the same identical name. But I want to keep the lowest dilution to be the one that gets compared to ESF data. 
df_picarro2 <- df_picarro %>%
  group_by(Name) %>%
  arrange(desc(Dilution), .by_group = TRUE) %>% # highest dilution first
  mutate(
    n = n(), # of duplicates
    idx = row_number(),  # 1 = highest dilution
    Name = case_when(
      n == 1 ~ Name, # only one row = keep name
      idx == n ~ Name,  # lowest dilution = keep name
      TRUE ~ paste0(Name, idx))) %>%# others =add suffix
    ungroup()

df_picarro2$Name[df_picarro2$Name == 'Abyssal_out_NW'] <- 'Abyssal_in_NW' #in/out doesn't really exist, so for having common names across sites, changing this to in. 
```
#merge datasets
```{r}
#LiCOR methane + Picarro methane/isotopes merged together
#dataframe keeping ALL rows from both datasets
#mergedall <- merge(pocks, df_picarro2, by = "Name", all=TRUE)
#mergedall$VialC <-as.numeric(mergedall$VialC)
#mergedall$VialC_ppm <- (mergedall$VialC/1000)
#mergedall$difference <- mergedall$CH4.Conc - mergedall$VialC_ppm
#write.csv(mergedall, "mergedall2.csv", row.names = FALSE)

#df only keeping rows where Name is the same across both datasets (removes zero/3 ppms, BigTank etc= only pockmarks)
merged <- merge(pocks, df_picarro2, by = "Name")
merged$C_wat0 <-as.numeric(merged$C_wat0)
merged$VialC <-as.numeric(merged$VialC)
merged$VialC_ppm <- (merged$VialC/1000)
merged$CH4.Conc <-as.numeric(merged$CH4.Conc)
merged$difference <- merged$CH4.Conc - merged$VialC_ppm
merged$abs_diff <- abs(merged$difference)
```
#find outliers
```{r}
#Couple vials were bad (based on major sensor disagreement). Locate them and tag as outliers
# looking at absolute value of the difference
Q1 <- quantile(merged$abs_diff, 0.25, na.rm = TRUE)
Q3 <- quantile(merged$abs_diff, 0.75, na.rm = TRUE)
IQR <- Q3 - Q1
lower_bound <- Q1 - 3 * IQR
upper_bound <- Q3 + 3 * IQR

# flag it in the df
merged$OutlierFlag <- ifelse(merged$abs_diff < lower_bound | merged$abs_diff > upper_bound, "Outlier", "Normal")

# 12 Outliers Identified, but 2 are replicates (so 10 bad data points)
```
##Adjust outliers
```{r}
#Plot LiCOR methane as a function of Picarro Methane, and only look at data where there are no outliers (normal). Find the regression between the both, and everywhere the Licor data has an Outlier, correct that outlier value based on the regression slope and y. 
model <- lm(VialC_ppm ~ CH4.Conc, data = merged[merged$OutlierFlag == "Normal", ])
summary(model)

merged$CH4_corrected <- merged$CH4.Conc
merged$VialC_corrected <- merged$VialC_ppm

m  <- coef(model)[2]  # slope
b  <- coef(model)[1]  # intercept

# Extract slope and R2
coef(model)
modslope <- coef(model)[2]
modint <- coef(model)[1]
mod_r2 <- summary(model)$r.squared

#Correcting the values for the "outlier" samples where there was a large discrepancy between both instruments. 

for (i in seq_len(nrow(merged))) {
  if (merged$OutlierFlag[i] == "Outlier") {
    if (merged$CH4.Conc[i] > merged$VialC_ppm[i]) {
      merged$CH4_corrected[i] <- (merged$VialC_ppm[i] - b) / m} 
    else {
      merged$VialC_corrected[i] <- m * merged$CH4.Conc[i] + b}}}

#show only outliers
merged_short <-merged %>%
  filter(OutlierFlag=="Outlier")%>%
  dplyr::select(Name,CH4.Conc,VialC_ppm,CH4_corrected,VialC_corrected)

#show only clean dataframe
merged_clean <-merged %>%
  dplyr::filter(OutlierFlag != "Outlier")
  
ggplot(merged_clean, aes(x =CH4.Conc, y=VialC_ppm)) +
 geom_point() +
  geom_smooth(method = "lm",se = FALSE,color = "black")+
  theme_bw() +
  theme(axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5))+
  labs(x = expression(CH[4]~"flux (µmol m"^-2~"s"^-1~") Cornell"),
       y = expression(CH[4]~"flux (µmol m"^-2~"s"^-1~") ESF"))+
  annotate("text", 
        x = Inf, y = -Inf, hjust = 1.1, vjust = -0.5,
        label = paste0("Slope = ", round(modslope, 2),
                       "\nY-int = ", round(modint, 2),
                          "\nR² = ", round(mod_r2, 2)))

#write.csv(merged_short, "C:/Erin/Seneca_Drums/Outliers.csv")
#write.csv(merged, "C:/Erin/Seneca_Drums/CH4_full.csv")
```
#CWat0= CH4 dissolved
```{r}
#Cwat0 is the concentration that is in the water when we drew the sample (most derived of the vial): BEST TO USE= aquatic concentrations
#since methane from the licor was adjusted for some bad outliers with the clean vialC_corrected column, we need to make sure Cwat0 is also adjusted for those select vials that were "bad"

#recalculate Cwat0 using VialC_corrected

#keep constants
CH4 <- data.frame(kH = 0.0014, dlnkHdinvT = 1600)
pbar <- 991.15 * 0.001     # pressure bar
rho_wat <- 0.001           # Water density (kg mL-1)
R <- 0.0831446261815 * 1000  # gas constant mL bar K-1 mol-1
# Initial gas-phase concentration (ppm -> mol mL-1)
C_gas0 <- 3
C_gas0 <- C_gas0 * pbar / R / (21.11 + 273.15)#Assumed at ambient temp. Could be refined

#convert to numeric
merged$WAT_TEMP<-as.numeric(merged$WAT_TEMP)
merged$BOAT_AIR_VOL<-as.numeric(merged$BOAT_AIR_VOL)
merged$BOAT_WAT_VOL<-as.numeric(merged$BOAT_WAT_VOL)

#temperature corrected Henrys constant
merged$kH <- CH4$kH * exp(
  CH4$dlnkHdinvT *
    (1 / (merged$WAT_TEMP + 273.15) - 1 / 298.15))

#original calculatations, using VialC (not yet converted to ppm) <- <- just confirming math here is the same as original calculations done by Tim. 
#data$WatC1<-data$VialC/(10^9)*pbar*kH*rho_wat #mol/mL
merged$WatC1_raw <- (merged$VialC / 10^9) *
                    pbar * merged$kH * rho_wat #mol/mL

#data$GasC1<-data$VialC*pbar/R/(data$WAT_TEMP+273.15)/(10^9) #mol/mL
merged$GasC1_raw <- merged$VialC * pbar / R /
                    (merged$WAT_TEMP + 273.15) / (10^9)

merged$cWat_0_raw=(merged$GasC1_raw*merged$BOAT_AIR_VOL+merged$WatC1_raw*-C_gas0*merged$BOAT_AIR_VOL)/merged$BOAT_WAT_VOL #mol mL-1

#using the clean, fixed VialC_corrected
#divide by 1e6 not 1e9 because VialC corrected is in ppm now. 
merged$WatC1_corr <- (merged$VialC_corrected / 1e6) *
                     pbar * merged$kH * rho_wat

merged$GasC1_corr <- merged$VialC_corrected * pbar / R /
                     (merged$WAT_TEMP + 273.15) / 1e6

#variable of cWat_0 but with fixed data. 
merged$cWat_0_corr <- (merged$GasC1_corr*merged$BOAT_AIR_VOL+merged$WatC1_corr*-C_gas0*merged$BOAT_AIR_VOL)/merged$BOAT_WAT_VOL

merged$cWat_0_corr2<-merged$cWat_0_corr * 1e9 #nmol

colnames (merged)
#trim down to have tidy datafrme with important variables. 
merged2<- merged %>%
  dplyr::select(Date, Name, Site_Num, Pockmark, in.out, Site, Depth, 
         OutlierFlag, CH4.Conc,CH4_corrected,VialC_ppm, 
         VialC_corrected,C_wat0,cWat_0_corr,cWat_0_corr2)
```
#Add DEC Data
```{r}
#Date EXO2 data (python-output df) and the chloride water samples to the methane dateframe
library(lubridate)
clbr <- read.csv("Cl_Br.csv") #chloride, only Pockmark IN
python <- read.csv("PythonRv2.csv") #EXO2 sensor data
python <- python %>%
  filter(Comments != "TEST") #remove test data
colnames(python)

#remove blanks, trim down to only important columns
clbr <- clbr %>%
  filter(!grepl("Blank", SiteName))%>%
  dplyr::select(Name, Cl_mgL)

#last 5 samples taken from each EXO2 measurement and then averaged
python5 <- python %>%
  mutate(IN_OUT = tolower(IN_OUT),
         datetime = mdy(Date) + hms(Time))  %>%
  group_by(SiteName, IN_OUT, Cardinal) %>%
  filter(!grepl("equipblank1", SiteName))%>%
  filter(!grepl("equipblank2", SiteName))%>%
  arrange(datetime, .by_group = TRUE) %>%
  slice_tail(n = 5) %>%
  summarise(
    across(
      where(is.numeric),
      ~ mean(.x, na.rm = TRUE)),
    .groups = "drop") %>%
  mutate(
    Name = paste(SiteName, IN_OUT, Cardinal, sep = "_")) %>%
 dplyr:: select(SiteName,Name, IN_OUT, Cond, ODOpercLocal, 
                Press,Baro,pH,pHmv,Turb,TDS,SpCond,Sal)


merged3 <- merge(merged2, python5, by = "Name", all=TRUE)
merged4 <- merge(merged3, clbr, by = "Name", all=TRUE)
all <- merged4
all <- all[!is.na(all$Pockmark), ]
all <- all[!is.na(all$Cond), ]

#write.csv(all, "C:/Erin/Seneca_Drums/all_data_clean.csv")
```
#CH4 STATS
```{r}
library(car) #for model ANOVAs
library(performance) #performance check
library(DHARMa)  #performance check
library(glmmTMB) #for fixed/random effects in models

#we want model comparisons to be based on the mean across all factors, not the first level (alphabetical)
options("contrasts")
options(constrasts=c("contr.sum","contr.poly"))  #compares to  global mean rather than a reference variable
#contrasts(all$Pockmark) <- contr.sum(levels(all$Pockmark))

all$Site_Num<-as.factor(all$Site_Num)
all$Site_Num <- factor(all$Site_Num, levels = c("A", "B", "1","2","3","4","5","6","7","8","9","10","11","12","13","14"))
levels(all$Site_Num)

all <- all %>%
  mutate(sample_id = paste(Site_Num, in.out, sep = "_"))

#df with no controls (because these don't have In vs Out)
treatments<- all  %>%
  filter(!Pockmark=="Abyssal")%>%
  filter(!Pockmark=="Belhurst") %>%
  mutate(sample_id = paste(Site_Num, in.out, sep = "_"))

#make column with control vs treatment
all <- all %>%
  mutate(cont_treat = case_when(
    grepl("^(Abyssal_|Belhurst_)", Name) ~ "Control", TRUE~ "Treatment"))

#make column with control vs each Pockmark Name retained
#all <- all %>% mutate(Pockmark3 = case_when(grepl("^(Abyssal_|Belhurst_)", Name) ~ "Control",TRUE ~ Pockmark))

all$sample_id<-as.factor(all$sample_id)
all$sample_id <- factor(all$sample_id, levels = c("A_in","A_out", "B_in","B_out", "1_in","1_out", "2_in", "2_out", "3_in","3_out","4_in","4_out","5_in","5_out","6_in","6_out","7_in","7_out","8_in","8_out","9_in","9_out","10_in","10_out","11_in", "11_out","12_in","12_out","13_in","13_out","14_in","14_out"))

levels(all$Site_Num)
levels(all$sample_id)
all$cont_treat<-as.factor(all$cont_treat)

#all_in <- all %>%  filter(in.out =="in")
#all_out <- all %>% filter(in.out =="out")

```
##ch4 ~ all sites
```{r}
library(multcompView) 
library(multcomp) #cld letters for plots
library(emmeans) #pairwise comparisons

plot(all$cWat_0_corr)

#random effect of Site_num/sample_id to account for replicates. This accounts for 1) overall site differences andd models in.out variablility. Wheras site_num:sample_id does not do that. 
m_control <- glmmTMB(
cWat_0_corr2 ~ Site_Num + (1 | Site_Num/sample_id), 
#to inform the replicates I have per sample
family = Gamma(link = "log"),
data = all)
summary(m_control)
car::Anova(m_control)
#check_model(m_control)
#simulate_residuals(m_control)
pock_lett<-emmeans(m_control, ~Site_Num, type="response") %>%
multcomp::cld(Letters = "abcdefghijk")

```
##pockmark * in_out
```{r}
#Treatments ONLY (true pockmarks, not controls)

#continuous data that is skewed.Tried tweedie= did not converge.Gaussian, also bad fit, indicated by residuals.Controls (belhurst/abyssal excluded from this analysis because they don't have in/out which would cause singularity)  
test<-glmmTMB(
  cWat_0_corr2 ~ Site_Num * in.out + (1 | Site_Num/in.out),
  family = Gamma(link="log"),
  data = treatments) #excludes Abyssal and Belhurst
summary(test)
car::Anova(test, type=3)
#check_model(test) #VIF OK because of nesting structure. 
#simulateResiduals(test, plot=TRUE)

meth_abc<-emmeans(test, ~Site_Num, type="response") %>% 
multcomp::cld(Letters = "abcdefghijk")

meth_in<-emmeans(test, ~in.out|Site_Num, type="response") %>%
multcomp::cld(Letters = "abcdefghijk")

test2<-glmmTMB(
  cWat_0_corr2 ~ in.out + (1 | Site_Num/in.out),
  family = Gamma(link="log"),
  data = treatments)
summary(test2) #borderline 
car::Anova(test2, type=3)
#simulateResiduals(test2, plot=TRUE)
meth_in.out<-emmeans(test2, ~in.out, type="response") %>% 
multcomp::cld(Letters = "abcd")
```
##control vs treat
```{r}
#controls 
m_control_mixed <- glmmTMB(
  cWat_0_corr2 ~ cont_treat + (1 | sample_id), 
  family = Gamma(link="log"), 
  data = all)
summary(m_control_mixed)
#check_model(m_control_mixed)
#simulateResiduals(m_control_mixed, plot=TRUE)

#no differnce between controls/treatment
car::Anova(m_control_mixed, type=3)
c_t_letters<-emmeans(m_control_mixed, ~cont_treat, type = "response") %>% 
  multcomp::cld(Letters = "abcdefghijk")

#emmeans(m_control_mixed, ~cont_treat, type = "response") %>%contrast("trt.vs.ctrl", ref = "Control")

#check_model(m_control_mixed)
```
##CH4 ~exo2 mods
```{r}
#checking to see whether methane varied by other environmental variables, both as a whole and by Site specifically

#We used linear mixed-effects models to assess the influence of environmental variables on methane fluxes while accounting for non-independence of observations within sites. Site was included as a random intercept, and environmental variables were treated as fixed effects. Methane fluxes were log-transformed to meet model assumptions.

all$log_CH4 <- log(all$cWat_0_corr + 0.001)

library(lme4)

#1) model doesn't converge when using glmmTMB for random effects --> using lmer; 2) model doesn't converge with log_CH4 as y, or with sample_id as fixed effect. 3) when (1 | Site_Num/sample_id) = boundary (singular) fit: see help('isSingular'). 


#TDS and SpCond = correlated. Removing TDS

test_m1 <- lmer(cWat_0_corr2 ~ Depth+ SpCond+ODOpercLocal+Turb+ (1 | Site_Num),data = all)
summary(test_m1)
car::Anova(test_m1, type=3)
#check_model(test_m1)

#log transforming does not change much
test_m2 <- lmer(log_CH4 ~ Depth+ SpCond+ODOpercLocal+Turb+ (1 | Site_Num),data = all)
summary(test_m2)
car::Anova(test_m2, type=3) #same as original test 1
#check_model(test_m2)

#scaling since all the variables have wildly different unit sizes. 
all <- all |>
  dplyr::mutate(
    Depth_s   = scale(Depth),
    SpCond_s  = scale(SpCond),
    ODOperc_s = scale(ODOpercLocal),
    TDS_s     = scale(TDS), #don't need
    Turb_s    = scale(Turb))

test_m3 <- lmer(cWat_0_corr2 ~ Depth_s+ SpCond_s+ODOperc_s+Turb_s+ (1 | Site_Num),data = all)
summary(test_m3)
car::Anova(test_m3, type=3)
#check_model(test_m3)

test_m4 <- lmer(log_CH4 ~ Depth_s+ SpCond_s+ODOperc_s+Turb_s+ (1 | Site_Num), data = all)
summary(test_m4)
car::Anova(test_m4, type=3)

test_m5 <- lm(log_CH4 ~ Depth_s+ SpCond_s+ODOperc_s+Turb_s, data = all)
summary(test_m5)

#test 1 good. Other variations didn't improve, so sticking with the one that doesn't transform the data

```
###Turbidity outlier
```{r}
#the one turbidity point thhat was really high masked the significance of DO! Removed The Turb outlier (> 3 STD DEV)
turb_mean <- mean(all$Turb, na.rm = TRUE)
turb_sd   <- sd(all$Turb, na.rm = TRUE)

# Flag outliers: TRUE if more than 3 SD from mean
all$Turb_outlier <- abs(all$Turb - turb_mean) > 3 * turb_sd

# Check how many outliers
table(all$Turb_outlier)
all_clean <- all[!all$Turb_outlier, ]

#Table 2
test_m6 <- lmer(cWat_0_corr2 ~ Depth+ SpCond+ODOpercLocal+Turb+ (1 | Site_Num),data = all_clean)
summary(test_m6)
car::Anova(test_m6, type=3)
#check_model(test_m6)
```
##CH4 Plots
###CH4~ EXO2/Cl
```{r}
library(ggeffects)
#plot(ggpredict(test_m6, terms = "ODOpercLocal"))
library(ggpubr) #ggarrange plots
library(MuMIn)
# cWat_0_corr2 = n mol / mL

#all_in <- all %>% filter(in.out=="in")
#max(all_in$cWat_0_corr2) #1.024362e-10
#min(all_in$cWat_0_corr2) #4.656343e-11

#all_out <- all %>% filter(in.out=="out")
#max(all_out$cWat_0_corr2) #1.024362e-10
#min(all_out$cWat_0_corr2) #4.656343e-11

#max(all$cWat_0_corr2) #1.024362e-10
#min(all$cWat_0_corr2) #4.656343e-11


#same additive model as above, just moved again here for clarity
test_m6 <- lmer(cWat_0_corr2 ~ Depth+ SpCond+ODOpercLocal+Turb+ (1 | Site_Num),data = all_clean)
summary(test_m6)
car::Anova(test_m6, type=3)

r2_values <- r.squaredGLMM(test_m6)
r2_marginal <- r2_values[1]
r2_conditional <- r2_values[2]

# Fixed effects
coefs <- fixef(test_m6)

# 2. DEPTH
depth_eff <- ggpredict(test_m6, terms = "Depth")
m_depth <- coefs["Depth"]
b_depth <- coefs["(Intercept)"]


CH4_depth <- ggplot() +
  geom_point(data = all_clean,
             aes(x = Depth, y = (cWat_0_corr2)),
             alpha = 0.6) +
  geom_line(data = depth_eff,
            aes(x = x, y = predicted),
            color = "black", size = 1) +
  geom_ribbon(data = depth_eff,
              aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.2) +
  theme_bw() +
  theme(text = element_text(size = 14, family = "serif")) +
  labs(x = "Depth (m)",
       y = expression(CH[4]~(nmol~mL^-1))) +
  annotate("text",
           x = max(depth_eff$x),
           y = max(depth_eff$predicted)+0.012,
           hjust = 1,
           label = paste0("Slope = ", signif(m_depth, 3),
                          "\nIntercept = ", signif(b_depth, 3),
                          "\nMarginal R² = ", round(r2_marginal, 2)))


# 3. SPECIFIC CONDUCTIVITY

sp_eff <- ggpredict(test_m6, terms = "SpCond")
m_sp <- coefs["SpCond"]
b_sp <- coefs["(Intercept)"]

CH4_spcond <- ggplot() +
  geom_point(data = all_clean,
             aes(x = SpCond, y = cWat_0_corr2),
             alpha = 0.6) +
  geom_line(data = sp_eff,
            aes(x = x, y = predicted),
            color = "black", size = 1) +
  geom_ribbon(data = sp_eff,
              aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.2) +
  theme_bw() +
  theme(text = element_text(size = 14, family = "serif")) +
  labs(x =  expression(Specific~conductivity~(µg~L^-1)),
        # "Specific conductivity (µg/L)",
       y = expression(CH[4]~(nmol~mL^-1))) +
  annotate("text",
           x = max(sp_eff$x),
           y = max(sp_eff$predicted)+0.012,
           hjust = 1,
           label = paste0("Slope = ", round(m_sp, 4),
                          "\nIntercept = ", signif(b_sp, 3),
                          "\nMarginal R² = ", round(r2_marginal, 2)))


# 4. DISSOLVED OXYGEN
do_eff <- ggpredict(test_m6, terms = "ODOpercLocal")
m_do <- coefs["ODOpercLocal"]
b_do <- coefs["(Intercept)"]

CH4_DO <- ggplot() +
  geom_point(data = all_clean,
             aes(x = ODOpercLocal, y = cWat_0_corr2),
             alpha = 0.6) +
  geom_line(data = do_eff,
            aes(x = x, y = predicted),
            color = "black", size = 1) +
  geom_ribbon(data = do_eff,
              aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.2) +
  theme_bw() +
  theme(text = element_text(size = 14, family = "serif")) +
  labs(x = "Dissolved Oxygen (%)",
       y = expression(CH[4]~(nmol~mL^-1))) +
  annotate("text",
           x = max(do_eff$x),
           y = max(do_eff$predicted)+0.012,
           hjust = 1,
           label = paste0("Slope = ", round(m_do, 4),
                          "\nIntercept = ", signif(b_do, 3),
                          "\nMarginal R² = ", round(r2_marginal, 2)))


# 5. TURBIDITY

turb_eff <- ggpredict(test_m6, terms = "Turb")

m_turb <- coefs["Turb"]
b_turb <- coefs["(Intercept)"]

CH4_Turb <- ggplot() +
  geom_point(data = all_clean,
             aes(x = Turb, y = cWat_0_corr2),
             alpha = 0.6) +
  geom_line(data = turb_eff,
            aes(x = x, y = predicted),
            color = "black", size = 1) +
  geom_ribbon(data = turb_eff,
              aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.2) +
  theme_bw() +
  theme(text = element_text(size = 14, family = "serif")) +
  labs(x = "Turbidity (NFU)",
       y = expression(CH[4]~(nmol~mL^-1))) +
  annotate("text",
           x = max(turb_eff$x),
           y = max(turb_eff$predicted)+0.012,
           hjust = 1,
           label = paste0("Slope = ", round(m_turb, 4),
                          "\nIntercept = ", signif(b_turb, 3),
                          "\nMarginal R² = ", round(r2_marginal, 2)))

ggarrange(CH4_depth,
          CH4_DO,
          CH4_spcond,
          CH4_Turb,
          ncol = 2,
          nrow = 2,
          labels = "auto")

```
###CH4 ~ Pockmarks
```{r}
library(ggpubr)
#to ease interpretation, we changed the plot units to nmol/ mL instead of mol/ mL; however, the stats (sig letters) are based on the original data (mol/ML)

dodge_width=0.7

#methane in nm is used for the plot, but stats comes from meth_in model which is mol. 
meth_in2 <- treatments %>%
  group_by(Site_Num, in.out) %>%
  summarise(y_max = max(cWat_0_corr2, na.rm = TRUE), .groups = "drop") %>%
  left_join(meth_in, by = c("Site_Num", "in.out")) %>%
  mutate(y_pos = y_max * 1.05) 

p1<-ggplot(treatments, aes(x=Site_Num, y=cWat_0_corr2, fill=in.out)) +
  geom_boxplot(alpha=0.7) +
 # facet_grid(~Pockmark)+
  #geom_jitter( alpha=0.3) +
  labs(x = "",
      y = expression(CH[4]~(nmol~mL^-1)))+
    geom_text(
    data = meth_in2,
    aes(x = Site_Num, y = y_pos, label = .group),
    position = position_dodge(width = dodge_width),
    size = 3.5)+
 # geom_text_repel(aes(label = meth_in), vjust = -0.5)+ 
    #nudge_y=.1)+
      # title="CH4 concentrations: in vs out") +
  scale_fill_discrete(labels = c("in" = "In",
                                 "out" = "Out"))+
  theme_bw()+
  theme(text = element_text(size = 12,family = "serif"),
        axis.text.x = element_text(hjust=1),
        legend.title=element_blank(),
        legend.position = "right")

c_t_letters2 <- all %>%
  group_by(cont_treat) %>%
  summarise(y_max = max(cWat_0_corr2, na.rm = TRUE), .groups = "drop") %>%
  left_join(c_t_letters, by = c("cont_treat")) %>%
  mutate(y_pos = y_max * 1.05) 


p2<-ggplot(all, aes(x=cont_treat, y=cWat_0_corr2)) +
  geom_boxplot(alpha=0.7) +
 # facet_grid(~Pockmark)+
  geom_jitter(width=0.15, alpha=0.3) +
 # scale_y_continuous(trans="log10") + 
  labs(x = "",
    y = expression(CH[4]~(nmol~mL^-1)))+
   geom_text(
    data = c_t_letters2,
    aes(x = cont_treat, y = y_pos, label = .group),
    position = position_dodge(width = dodge_width),
    size = 3.5)+
      # title="CH4 concentrations: in vs out") +
  theme_bw()+
  theme(text = element_text(size = 12,family = "serif"),
        legend.title=element_blank(),
        legend.text = element_blank(),
        legend.position = "none")


meth_in.out2 <- treatments %>%
  group_by(in.out) %>%
  summarise(y_max = max(cWat_0_corr2, na.rm = TRUE), .groups = "drop") %>%
  left_join(meth_in.out, by = c("in.out")) %>%
  mutate(y_pos = y_max * 1.05) 


p3<-ggplot(treatments, aes(x=in.out, y=cWat_0_corr2)) +
  geom_boxplot(alpha=0.7) +
  #stat_summary(fun.y = mean, geom = "point", shape = 18, size = 3, color = "darkred")+
  #geom_jitter() +
 labs( x="",y = expression(CH[4]~(nmol~mL^-1)))+
    geom_text(data = meth_in.out2, size = 3.5,
    aes(x = in.out, y = y_pos, label = .group))+
  scale_x_discrete(labels = c("in" = "In",
                              "out" = "Out"))+
      # title="CH4 concentrations: in vs out") +
  theme_bw() +
  theme(text = element_text(size = 12,family = "serif"),
        legend.title=element_blank(),
        legend.position = "none")


pock_lett2 <- all %>%
  group_by(Site_Num) %>%
  summarise(y_max = max(cWat_0_corr2, na.rm = TRUE), .groups = "drop") %>%
  left_join(pock_lett, by = c("Site_Num")) %>%
  mutate(y_pos = y_max * 1.05) 

p4<-ggplot(all, aes(x=Site_Num, y=cWat_0_corr2)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
 # scale_y_continuous(trans="log10") + 
 labs( x="", y = expression(CH[4]~(nmol~mL^-1)))+
    geom_text(
    data = pock_lett2,
    aes(x = Site_Num, y = y_pos, label = .group),
    size = 3.5)+
    #vjust = -.5,
   # nudge_y=.1)+
      # title="CH4 concentrations by Pockmark") +
  theme_bw() +
  theme(text = element_text(size = 12,family = "serif"),
    #axis.text.x = element_text( hjust=1),
        axis.title.x=element_blank(),
        plot.title=element_text(hjust=0.5),
    legend.position = "none")

gg<-ggarrange(p2,p3, legend="none")
ggarrange(p4, p1, p2, p3 ,nrow=2, ncol=2, align="hv", labels="auto")
```
#Chloride/SpCond
##spcond= proxy
```{r}
#is Specific conductivity a proxy for Chloride?
plot(all$SpCond)
cl_cond <- lm(SpCond ~ Cl_mgL, data=all)
summary(cl_cond)
shapiro.test(residuals(cl_cond)) #0.02, not normal
#cor.test(all$Cl_mgL, all$SpCond, method = "spearman")
#plot(cl_cond)
#check_model(cl_cond)
#simulateResiduals(cl_cond, plot=TRUE)
m_clsp  <- coef(cl_cond)[2]  # slope
b_clsp  <- coef(cl_cond)[1]  # intercept
r2_clsp <- summary(cl_cond)$r.squared

#specific conductivity can be a proxy. Use this instead
cond_cl_p<-ggplot(all, aes(x=Cl_mgL, y=SpCond)) +
 geom_point() +
  geom_smooth(method = "lm", se = TRUE,color = "black")+
  ylim(630,725)+
  theme_bw() +
  theme(text = element_text(size = 12,family = "serif"),
        axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5))+
  labs(x =  expression(Chloride~(mg~L^-1)),
   # x = "Chloride (mg/L)",
      y =  expression(Specific~conductivity~(µg~L^-1)))+
      # y = "Specific Conductivity (µg/L)")+
  annotate("text", 
        x = 142, y = 650, hjust = 1.1, vjust = -0.5,
        label = paste0("Slope = ", round(m_clsp, 2),
                       "\nY-int = ", round(b_clsp, 2),
                          "\nR² = ", round(r2_clsp, 2)))
```
##Spcond ~all  site
```{r}
#does specific conductivity vary by all sites?
SpCond_site <- glmmTMB(
  SpCond ~ Site_Num +(1|sample_id),
  family = Gamma(link = "log"),
  data = all)
summary(SpCond_site)
car::Anova(SpCond_site)
#check_model(SpCond_site) #great pp plot
#simulate_residuals(SpCond_site)
sp_site_lett<-emmeans(SpCond_site, ~Site_Num, type="response") %>%
multcomp::cld(Letters = "abcdefghijk")

#sp cond by all sites
c1<-ggplot(all, aes(x=Site_Num, y=SpCond)) +
  geom_boxplot(alpha=0.7) +
    geom_text(
    data = sp_site_lett,
    aes(x = Site_Num, y = (response+12), label = .group),
    size = 3)+
  #geom_jitter()+
  labs( x="All Sites", 
         y =  expression(Specific~conductivity~(µg~L^-1)))+
  #y = "Specific Conductivity (µg/L)") +
  ylim(600,750)+ #removes 2 Belhurst points
  theme_bw()+
  theme(text = element_text(size = 12,family = "serif"))
```
## spcond ~ in/out
```{r}
#sp cond by in or out (treatments only)
#convergence issues, cannot use interaction effect
SpCondmod2 <- glmmTMB(
  SpCond ~ Site_Num*in.out,
  family = Gamma(link = "log"),
  data = treatments)

#Use this one
SpCondmod3 <- glmmTMB(
  SpCond ~ Site_Num+in.out +(1| sample_id),
  family = Gamma(link = "log"),
  data = treatments)
summary(SpCondmod3)
car::Anova(SpCondmod3)
check_model(SpCondmod3)

spcond_lett3<-emmeans(SpCondmod3, ~Site_Num, type="response") %>%
multcomp::cld(Letters = "abcdefghijk")


dodge_width <- 0.75 # same as default for geom_boxplot
c1b<-ggplot(treatments, aes(x = Site_Num, y = SpCond, fill=in.out))+
  geom_boxplot(alpha = 0.7) +
  geom_jitter(
    width = 0.15,
    size = 1.5,
    alpha = 0.8) +
  ylim(620,700)+
  labs(x="Pockmarks", 
       y =  expression(Specific~conductivity~(µg~L^-1)))+
  #y = "Specific Conductivity (µg/L)")+
 # geom_text(
 # data = spcond_lett2,
 # aes(x = Site_Num, y = 698, label = .group),
  #position = position_dodge(width = dodge_width),
 # size = 3)+
   scale_fill_discrete(labels = c("in" = "In",
                                 "out" = "Out"))+
  theme_bw() +
  theme(text = element_text(size = 12,family = "serif"),
    #axis.text.x = element_text(angle = 45, hjust = 1),
   # axis.title.x= element_blank(),
    legend.title = element_blank())
```
##spcond ~depth
```{r}
#sp cond by depth
cl_depth <- lm(SpCond ~ Depth, data = all)
summary(cl_depth)
m_dep  <- coef(cl_depth)[2]  # slope
b_dep  <- coef(cl_depth)[1]  # intercept
r2_dep <- summary(cl_depth)$r.squared

c2<-ggplot(all, aes(x =Depth, y=SpCond)) +
 geom_point() +
  geom_smooth(method = "lm", se = TRUE,color = "black")+
  ylim(600,721)+
  theme_bw() +
  theme(text = element_text(size = 12,family = "serif"),
        axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5))+
  labs(x = "Depth (m)",
       y =  expression(Specific~conductivity~(µg~L^-1)))+
  annotate("text", 
        x = -50, y = 675, hjust = 1.1, vjust = -0.5,
        label = paste0("Slope = ", round(m_dep, 2),
                       "\nY-int = ", round(b_dep, 2),
                          "\nR² = ", round(r2_dep, 2)))

```
#manuscript figs/tables
```{r}
#ggarrange(cond_cl_p,c3,c2,c1, nrow=2, ncol=2, labels="AUTO")

#Figure 1: site map
#Figure 2: cross section of lake
#Figure 3
Figure3<- ggarrange(p4, p1, p2, p3 ,nrow=2, ncol=2, align="hv", labels="auto")
ggsave(
  "Figure3.tiff",
  plot = Figure3,
  width = 3200,
  height = 2000,
  units = "px",
  dpi = 300,
  compression = "lzw")

#Figure 4
Figure4<-ggarrange(CH4_depth,CH4_DO,CH4_spcond, CH4_Turb, ncol=2, nrow=2, labels="auto")
ggsave(
  "Figure4.tiff",
  plot = Figure4,
  width = 3200,
  height = 2000,
  units = "px",
  dpi = 300,
  compression = "lzw")

#Figure 5
Figure5<-ggarrange(cond_cl_p,c2,c1,c1b, nrow=2, ncol=2, labels="auto", align="hv")
ggsave(
  "Figure5.tiff",
  plot = Figure5,
  width = 3200,
  height = 2000,
  units = "px",
  dpi = 300,
  compression = "lzw")

#figure6; hypothetical examples

#Table 1
test<-glmmTMB(
  cWat_0_corr ~ Site_Num * in.out + (1 | Site_Num/in.out),
  family = Gamma(link="log"),
  data = treatments) #excludes Abyssal and Belhurst
summary(test)
car::Anova(test, type=3)

#Table 2
test_m6 <- lmer(cWat_0_corr ~ Depth+ SpCond+ODOpercLocal+Turb+ (1 | Site_Num),data = all_clean)
summary(test_m6)
car::Anova(test_m6, type=3)
```
#Other EXO2 vars
##stats
```{r}
#Normality not tested for at the moment! ANOVA (AOV) may not be appropriate 
get_letters_anova <- function(df, response, y_offset = NULL) {
  # build formula
  f <- as.formula(paste(response, "~ Site_Num"))
  # fit model
  m <- aov(f, data = df)
  # estimated marginal means + letters
  cld_df <- cld(
    emmeans(m, ~ Site_Num),
    Letters = letters,
    adjust = "tukey") %>%
    as.data.frame()
  # default y offset = 5% of data range
  if (is.null(y_offset)) {
    y_offset <- 0.05 * diff(range(df[[response]], na.rm = TRUE))}
  # add y-position for plotting
  cld_df <- cld_df %>%
    left_join(
      df %>%
        group_by(Site_Num) %>%
        summarise(
          y = max(.data[[response]], na.rm = TRUE) +
            y_offset,
          .groups = "drop"),
      by = "Site_Num" )
  return(cld_df)
}

#depth_letters <- get_letters_anova(python5, "Depth")
tds_letters <- get_letters_anova(all, "TDS")
ph_letters   <- get_letters_anova(all, "pH")
do_letters <- get_letters_anova(all, "ODOpercLocal")
#cond_letters  <- get_letters_anova(python5, "Cond")
turb_letters  <- get_letters_anova(all, "Turb")
#cl_letters  <- get_letters_anova(python5, "Cl_mgL")

#Differences in environmental variables among pockmarks were tested using one-way analyses of variance (ANOVA) when assumptions of approximate normality were met, and Kruskal–Wallis tests followed by Dunn’s post hoc comparisons for non-normal variables. Significance groupings were visualized using compact letter displays.
```
##Plots
```{r}
str(all)
turb<-ggplot(all, aes(x=Site_Num, y=Turb)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
    geom_text(
    data = turb_letters,
    aes(x = Site_Num, y = y, label = .group),
    size = 3.5,
    vjust = 0)+
 # scale_y_continuous(trans="log10") + 
  labs(y="Turbidity (NFU)")+
      # title="CH4 concentrations by Pockmark") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        plot.title=element_text(hjust=0.5))

TDS<- ggplot(all, aes(x=Site_Num, y=TDS)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
    geom_text(
    data = tds_letters,
    aes(x = Site_Num, y = y, label = .group),
    size = 3.5,
    vjust = 0)+
 # scale_y_continuous(trans="log10") + 
  labs(y="TDS")+
      # title="CH4 concentrations by Pockmark") +
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        plot.title=element_text(hjust=0.5))

pH<- ggplot(all, aes(x=Site_Num, y=pH)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
 # ylim(-90,-55)+
    geom_text(
    data = ph_letters,
    aes(x = Site_Num, y = y, label = .group),
    size = 3.5,
    vjust = 0)+
 # scale_y_continuous(trans="log10") + 
  labs(y="pH")+
      # title="CH4 concentrations by Pockmark") +
  theme_bw() +
  theme(axis.title.x=element_blank(),
        plot.title=element_text(hjust=0.5))

DO<- ggplot(all, aes(x=Site_Num, y=ODOpercLocal)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
    geom_text(
    data = do_letters,
    aes(x = Site_Num, y = y, label = .group),
    size = 3.5,
    vjust = 0)+
 # scale_y_continuous(trans="log10") + 
  labs(y="DO (%)")+
      # title="CH4 concentrations by Pockmark") +
  theme_bw() +
  theme(axis.title.x=element_blank(),
        plot.title=element_text(hjust=0.5))

ggarrange(turb, TDS, DO, pH, nrow=2,ncol=2)

```
##in/out
```{r}
turb2<-ggplot(all, aes(x=Site_Num, y=Turb, fill=in.out)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
  labs(y="Turbidity (NFU)")+
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        legend.title=element_blank())

TDS2<- ggplot(all, aes(x=Site_Num, y=TDS, fill=in.out)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
  labs(y="TDS (mg/L)")+
  theme_bw() +
  theme(axis.text.x = element_blank(),
        axis.title.x=element_blank(),
        legend.title=element_blank())

pH2<- ggplot(all, aes(x=Site_Num, y=pH, fill=in.out)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
 #ylim(-90,-55)+
  labs(y="pH")+
  theme_bw() +
  theme(#axis.text.x = element_text(angle=45, hjust=1),
        axis.title.x=element_blank(),
        plot.title=element_text(hjust=0.5),
        legend.title=element_blank())

DO2<- ggplot(all, aes(x=Site_Num, y=ODOpercLocal, fill=in.out)) +
  geom_boxplot(alpha=0.7) +
  geom_jitter(width=0.15, alpha=0.3) +
  labs(y="DO (%)")+
  theme_bw() +
  theme(#axis.text.x = element_text(angle=45,  hjust=1),
        axis.title.x=element_blank(),
        plot.title=element_text(hjust=0.5),
        legend.title=element_blank())

ggarrange( turb2, TDS2, DO2, pH2, nrow=2,ncol=2)
```
#UNUSED_________________
#with linear model
```{r}
break
#this is pulling linear model for each one

#methane ~ specific conductivity
ch4_spcond_mod <- lm(cWat_0_corr2 ~ SpCond, data = all)
summary(ch4_spcond_mod)
m_sp  <- coef(ch4_spcond_mod)[2]  # slope
b_sp  <- coef(ch4_spcond_mod)[1]  # intercept
ch_r2_sp <- summary(ch4_spcond_mod)$r.squared

CH4_spcond_p<-ggplot(all, aes(x=SpCond, y=cWat_0_corr2)) +
 geom_point() +
  geom_smooth(method = "lm", se = TRUE,color = "black")+
 # ylim(4.656343e-11,1.024362e-10)+
  #xlim(620,725)+
  theme_bw() +
  theme(text = element_text(size = 14,family = "serif"),
        axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5))+
  labs(x = "Specific conductivity (µg/L)",
     y = expression(CH[4]~(nmol~mL^-1)))+
  annotate("text", 
        x = 725, y = 0.09, hjust = 1.1,
        label = paste0("Slope = ", round(m_sp, 3),
                       "\nY-int = ", round(b_sp, 2),
                          "\nR² = ", round(ch_r2_sp, 2)))

#methane by Depth
ch4_depth <- lm(cWat_0_corr ~ Depth, data = all)
summary(ch4_depth)
m  <- coef(ch4_depth)[2]  # slope
b  <- coef(ch4_depth)[1]  # intercept
ch_r2 <- summary(ch4_depth)$r.squared

CH4_depth<-ggplot(all, aes(x=Depth, y=cWat_0_corr2)) +
 geom_point() +
  geom_smooth(method = "lm", se = TRUE,color = "black")+
  theme_bw() +
  theme(text = element_text(size = 14,family = "serif"),
        axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5))+
  labs(x = "Depth (m)",
     y = expression(CH[4]~(nmol~mL^-1)))+
  annotate("text", 
        x = -50, y = 0.08, hjust = 1.1, vjust = -0.5,
        label = paste0("Slope = ", round(m, 2),
                       "\nY-int = ", round(b, 2),
                          "\nR² = ", round(ch_r2, 2)))

#methane ~ DO
ch4_do <- lm(cWat_0_corr2 ~ ODOpercLocal, data = all)
summary(ch4_do)
m_do  <- coef(ch4_do)[2]  # slope
b_do<- coef(ch4_do)[1]  # intercept
do_r2 <- summary(ch4_do)$r.squared

DO<-ggplot(all, aes(x=ODOpercLocal, y=cWat_0_corr2)) +
 geom_point() +
  geom_smooth(method = "lm", se = TRUE,color = "black")+
  theme_bw() +
  theme(text = element_text(size = 14,family = "serif"),
        axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5))+
  labs(x = "Dissolved Oxygen (%)",
     y = expression(CH[4]~(nmol~mL^-1)))+
  annotate("text", 
        x = 102, y = 0.085, hjust = 1.1, vjust = -0.5,
        label = paste0("Slope = ", round(m_do,2),
                       "\nY-int = ", round(b_do, 2),
                          "\nR² = ", round(do_r2, 2)))
#methane ~ turb
ch4_turb <- lm(cWat_0_corr2 ~ Turb, data = all_clean)
summary(ch4_turb)
m_turb  <- coef(ch4_turb)[2]  # slope
b_turb <- coef(ch4_turb)[1]  # intercept
turb_r2 <- summary(ch4_turb)$r.squared

Turb_plot<-ggplot(all_clean, aes(x=Turb, y=cWat_0_corr2)) +
 geom_point() +
  geom_smooth(method = "lm", se = TRUE,color = "black")+
 # xlim(0,100)+
  theme_bw()+
  theme(text = element_text(size = 14,family = "serif"),
      axis.text.x = element_text(hjust = 1),
        plot.title = element_text(hjust=0.5))+
  labs(x = "Turbidity (NFU)",
     y = expression(CH[4]~(nmol~mL^-1)))+
  annotate("text", 
        x = 63, y = 0.065, hjust = 1.1, vjust = -0.5,
        label = paste0("Slope = ", round(m_turb,2),
                       "\nY-int = ", round(b_turb, 2),
                          "\nR² = ", round(turb_r2, 2)))


ggarrange(CH4_depth,DO,CH4_spcond_p, Turb_plot, ncol=2, nrow=2, labels="auto")
```